Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models

### 2015-05-14 tex/parametrized-learning.tex
new refs

While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameters of interest. In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$. When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers, so that one has $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.

This is a significant step forward in the usage of machine learning in HEP, because classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly relates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar is the so-called matrix element method in which one tries to directly calculate the target likelihood ratio by performing a computationally intensive integral associated to the detector response. The detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter.

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson and (if we are so lucky) measurement of the mass of supersymmetric particles in cascade decays.  Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects.

@@

We have shown how a generalized likelihood ratio test on a dataset of multiple i.i.d. events with several features can be reformulated in terms of the distributions of discriminative classifier output. This allows one to turn the optimization of the procedure into optimization of a per-event classification problem and a calibration procedure.

@@

This approach leverages the power of machine learning in a classical statistical setting.

>>>>

While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameters of interest. In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$. When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers. This is easily accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.

This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}. In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.  Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Bolognesi:2012mm}.

@@

We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional dataset with multiple events in terms of a univariate density of a classifier score. 
We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$
This allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure. This approach leverages the power of machine learning in a classical statistical setting.

@@
